Enhancing Student Engagement via Reduction of Frustration with Programming Assignments using Machine Learning

### 2017-02-14 paper.tex
First version new format, too long
Reduced to 14 pages without figures. Probably the state of the art can be cut by 75%

\documentclass[a4paper]{llncs}
\usepackage{subcaption} % an alternative package for sub figures
\usepackage{llncsdoc}

@@

% *** GRAPHICS RELATED PACKAGES ***
%
\ifCLASSINFOpdf
\usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are
\graphicspath{{./img/}}
  % and their extensions so you won't have to specify these with
  % every instance of \includegraphics
\DeclareGraphicsExtensions{.pdf,.jpeg,.png}
\else

\fi

@@

\author{Mario Garcia Valdez\inst{1}, Amaury Hernandez Aguila\inst{1}, Juan-J. Merelo\inst{2} and Alejandra Mancilla Soto\inst{1}}
\institute{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico)\\
    \email{\tt mario@tectijuana.edu.mx}
\and
Geneura Team and CITIC, University of Granada (Spain)\\
    \email{\tt jmerelo@ugr.es}
}

@@

Learning to program is often regarded as a difficult task, courses are 
problem-solving intensive and often students feel frustrated when doing programming
exercises. Instructors aim to design learning activities that result
challenging to students, but not to the extreme as to frustrate them. When
selecting an appropriate programing exercise, experienced instructors gage a
student’s affective state and skills to then assign an activity with the  appropriate level
of difficulty. This work is focused on the prediction of the affective states
of programmers with different levels of expertise when learning a new
programming language. For this, an interactive web-based programming platform
is proposed. The platform is designed to collect data from the students'
interaction for data analysis. Current work is focused on the prediction of
affective states using non-obtrusive sensors. Specifically,  the aim of this research 
is to evaluate the use of keyboard and mouse dynamics as an appropriate 
sensory input for an affective recognition
system. The proposed method uses a data mining approach in data 
generated from both keyboard and mouse dynamics of students as they 
work in programming assignments. The data set was generated
as students learned the basics of Python programming. Feature vectors 
were generated with 39 features representing the dynamics of the interaction. 
In total 149 feature vectors were generated and used as input of four different
classification algorithms: Na\"ive Bayes, J-48, k-NN, and Artificial Neural
Networks. These classification models are then compared by their performances
when classifying learners into five different affective states: boredom,
frustration, distraction, relaxation and engagement. Each of the classification
models achieved an accuracy of around 75\% with a kappa of around 0.45, and, in
average, J-48 was the better algorithm with an accuracy of 75.97\%. Results show
that the data gathered from the non-obtrusive sensors can successfully be used
as another input to classification models in order to predict an individual's
affective states.

\end{abstract}

@@

\begin{figure}[!t] 
\centering 
\includegraphics[width=2.5in]{KMDAffective.png} 
\caption{Code Execution and Affect Recognition Architecture.}
\label{fig_process} 
\end{figure}

@@

Although the use of Keystroke Dynamics (KD) is found in several research works
as a biometric measure, its use for identifying affective states is rare in
comparison. Epp, Lippold \& Mandryk \cite{epp2011identifying} effectively used KD in conjunction
with decision-tree classifiers for the identification of 15 affective states.
Although their work was based on fixed-text their technique to extract a feature
vector was an inspiration for the proposed method in this work. As for free-text
KD, Bixler y D'Mello \cite{bixler2013detecting} present a method for the identification of boredom
and engagement based on several classification models.

@@

\begin{table*}[!t]
\centering
\caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
\label{tab_features}
    \begin{tabular}{ | l | l || l | l | }
    \hline
    ID          & Description           & ID               & Description  \\
    \hline
    1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
    \hline
    2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
    \hline
    3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
    \hline
    4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
    \hline
    5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
    \hline
    6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
    \hline
    7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
    \hline
    8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
    \hline
    9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
    \hline
    10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
    \hline
    11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
    \hline
    12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
    \hline
    13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
    \hline
    14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
   \hline
    15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
    \hline
    16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
    \hline
    17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
    \hline
    18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
    \hline
    19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
    \hline
    20 & Standard deviation of the previous feature. &   &  \\
    \hline
  \end{tabular}
\end{table*}

@@

\begin{figure}[!t] 
\centering 
\includegraphics[width=3in]{editorRresult.png} 
\caption{Web-based editor used for code evaluation.}
\label{fig_editor} 
\end{figure}
\begin{figure}[b!] 
\centering 
\includegraphics[width=3in]{KeyDyn.png} 
\caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
\label{fig_KD} 
\end{figure}
\subsubsection{Preprocessing of the Keystroke and Mouse Data} 
The raw data
obtained from the script needs to be preprocessed to obtain a feature vector.
Basically, this pre-processing consists in measuring the delays between key-
down, key-up or mouse-move events triggered during an exercise. These events
have the dynamic shown in Figure 3; for example when typing the word ‘key’ a
user first presses the letter K and triggers the key-down event, this is
indicated with an arrow that changes the state of the key, the time of the event
is important and also recorded. Only the event data is received from the
browser, in order to generate feature vectors, the patterns and rhythm users
have when pressing consecutive keys are captured measuring the delays between
events, as it is a common practice when dealing with keystroke dynamics. In this
work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
time (Ht) is defined as the time between a key-down and a key-up of the same
key, this would be Ht(K) in the example. Inter key time (It) is defined as the
time between two consecutive key-down events, this time could be negative is the
second key is pressed before the first is released. A sequence is defined as a
list of consecutive keystrokes. In the above example valid sequences could be
[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
third is a trigraph. While there can be sequences of any size in a text, working
only with digraphs and trigraphs is preferred. Even if only digraphs and
trigraphs are considered, the amount found in free-text is too large and not
very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
they propose to use statistical measures to capture the patterns; the same
strategy is used in this work. The averages and standard deviations are
calculated from the delays between the events of the digraphs and trigraphs in
each program.

>>>>

\documentclass[a4paper]{llncs}

\usepackage{llncsdoc}

\usepackage{graphicx,grffile}

\graphicspath{{img/}}

@@

@@

\author{Mario Garcia Valdez\inst{1}, Amaury Hernandez Aguila\inst{1}, Juan-J. Merelo\inst{2} and Alejandra Mancilla Soto\inst{1}}
\institute{ Dept. of Graduate Studies, Instituto Tecnologico de Tijuana (Mexico),  \email{\tt \(mario,amherag\)@tectijuana.edu.mx}
\and
Geneura Team and CITIC, University of Granada (Spain), 
    \email{\tt jmerelo@ugr.es}
}

@@

Learning to program is often regarded as a difficult task. When
selecting an appropriate programing exercise, experienced instructors gauge a
student’s affective state and skills to then assign an activity with the  appropriate level
of difficulty. This work is focused on the prediction of the affective states
of programmers with different levels of expertise when learning a new
programming language. For this, an interactive web-based programming platform
is proposed. The platform is designed to collect data from the students'
interaction for data analysis. Current work is focused on the prediction of
affective states using non-obtrusive sensors. Specifically,  the aim of this research 
is to evaluate the use of keyboard and mouse dynamics as an appropriate 
sensory input for an affective recognition
system. The proposed method uses feature vectors obtained by mining  data 
generated from both keyboard and mouse dynamics of students as they 
work in basic Python programming assignments, which were used to train different
classification algorithms to classify learners into five different affective states: boredom,
frustration, distraction, relaxation and engagement. Accuracy achieved was around 75\% with J48 obtaining the best results, proving that data gathered from non-obtrusive sensors can successfully be used
as another input to classification models in order to predict an individual's
affective states.
\end{abstract}

@@

% \begin{figure}[!t] 
% \centering 
% \includegraphics[width=2.5in]{KMDAffective.png} 
% \caption{Code Execution and Affect Recognition Architecture.}
% \label{fig_process} 
% \end{figure}

@@

Although the use of Keystroke Dynamics (KD) is found in several research works
as a biometric measure, its use for identifying affective states is rare in
comparison. Epp, Lippold \& Mandryk \cite{epp2011identifying} effectively used KD in conjunction
with decision-tree classifiers for the identification of 15 affective states.
Although their work was based on fixed text, their technique to extract a feature
vector was an inspiration for the proposed method in this work. As for free-text
KD, Bixler y D'Mello \cite{bixler2013detecting} present a method for the identification of boredom
and engagement based on several classification models.

@@

% \begin{table*}[!t]
% \centering
% \caption{Table 1. The components of the feature vector; adapted from Epp, Lippold \& Mandryk \cite{epp2011identifying}.}
% \label{tab_features}
%     \begin{tabular}{ | l | l || l | l | }
%     \hline
%     ID          & Description           & ID               & Description  \\
%     \hline
%     1 & Mean duration between 1st and 2nd down keys of the digraphs.  & 21 & Mean duration of the 2nd key of the trigraphs.\\
%     \hline
%     2 & Standard deviation of the previous feature.                   & 22 & Standard deviation of the previous feature.\\
%     \hline
%     3 & Mean duration of the 1st key of the digraphs.                 & 23 & Mean duration between 2nd key up and next key down of trigraphs.\\
%     \hline
%     4 & Standard deviation of the previous feature.                   & 24 & Standard deviation of the previous feature.\\
%     \hline
%     5 & Mean duration between 1st key up and next key down of the digraphs. & 25 & Mean duration of the third key of the trigraphs.\\
%     \hline
%     6 & Standard deviation of the previous feature. & 26 & Standard deviation of the previous feature.\\
%     \hline
%     7 & Mean duration of the 2nd key of the digraphs. & 27 & Mean duration of the trigraphs from 1st key down to last key up.\\
%     \hline
%     8 & Standard deviation of the previous feature. & 28 & Standard deviation of the previous feature.\\
%     \hline
%     9 & Mean duration of the digraphs from 1st key down to last key up. & 29 & Mean number of key events that were part of the graph.\\
%     \hline
%     10 & Standard deviation of the previous feature. & 30 & Standard deviation of the previous feature.\\
%     \hline
%     11 & Mean number of key events that were part of the graph. & 31 & Mean duration of mouse key presses.\\
%     \hline
%     12 & Standard deviation of the previous feature. & 32 & Standard deviation of the previous feature.\\
%     \hline
%     13 & Mean duration between 1st and 2nd down keys of the trigraphs. & 33 & Mean duration of all mouse movements.\\
%     \hline
%     14 & Standard deviation of the previous feature. & 34 & Standard deviation of the previous feature.\\
%    \hline
%     15 & Mean duration of the 1st key of the trigraphs. & 35 & Mean distance in pixels in the X coordinate. \\
%     \hline
%     16 & Standard deviation of the previous feature. & 36 & Standard deviation of the previous feature.\\
%     \hline
%     17 & Mean duration between 1st key up and next key down of trigraphs. & 37 & Mean distance in pixels in the Y coordinate.\\
%     \hline
%     18 & Standard deviation of the previous feature. & 38 & Standard deviation of the previous feature.\\
%     \hline
%     19 & Mean duration between 2nd and 3rd down keys of the trigraphs. & 39 & Number of attempts\\
%     \hline
%     20 & Standard deviation of the previous feature. &   &  \\
%     \hline
%   \end{tabular}
% \end{table*}

@@

% \begin{figure}[!t] 
% \centering 
% \includegraphics[width=3in]{editorRresult.png} 
% \caption{Web-based editor used for code evaluation.}
% \label{fig_editor} 
% \end{figure}

% \begin{figure}[b!] 
% \centering 
% \includegraphics[width=3in]{KeyDyn.png} 
% \caption{Keystroke Dynamics, adapted from Sim \& Janakiraman \cite{sim2007digraphs}.}
% \label{fig_KD} 
% \end{figure}

\subsubsection{Preprocessing of the Keystroke and Mouse Data} 
The raw data
obtained from the script needs to be preprocessed to obtain a feature vector.
Basically, this pre-processing consists in measuring the delays between key-
down, key-up or mouse-move events triggered during an exercise. These events
have the dynamic shown in Figure 3; for example when typing the word ‘key’ a
user first presses the letter K and triggers the key-down event, this is
indicated with an arrow that changes the state of the key, the time of the event
is important and also recorded. Only the event data is received from the
browser, in order to generate feature vectors, the patterns and rhythm users
have when pressing consecutive keys are captured measuring the delays between
events, as it is a common practice when dealing with keystroke dynamics. In this
work, the definitions proposed by Sim y Janakiraman \cite{sim2007digraphs} are used. Held
time (Ht) is defined as the time between a key-down and a key-up of the same
key, this would be Ht(K) in the example. Inter key time (It) is defined as the
time between two consecutive key-down events, this time could be negative is the
second key is pressed before the first is released. A sequence is defined as a
list of consecutive keystrokes. In the above example valid sequences could be
[‘K’, ‘E’], [‘E’, ‘Y’] and [‘K’, ‘E’, ‘Y’], the first two are digraphs and the
third is a trigraph. While there can be sequences of any size in a text, working
only with digraphs and trigraphs is preferred. Even if only digraphs and
trigraphs are considered, the amount found in free-text is too large and not
very useful for a feature vector as noted by Epp, Lippold y Mandryk (2007) so
they propose to use statistical measures to capture the patterns; the same
strategy is used in this work. The averages and standard deviations are
calculated from the delays between the events of the digraphs and trigraphs in
each program.
